---
title: "Assignment 2"
author: "Your Name"
date: "Due January 26 by midnight"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```


## Assignment 2 Exercises

### Part 1

1 Look at the help file for the `mtcars` dataset. What is the data about?

2 Look at the structure of the `mtcars` data frame. How many observations are there? Variables? What do they represent?

3 Use the `summary` command on the `mtcars` data frame. What is being output?

4 Are all the variables truly numeric? Print the `cyl` variable. Then use the `unique` command on the `cyl` variable. What does this tell you?

5 Change the `cyl` variable to a factor. Check the levels of `cyl` after.

6 Get the subset of `mtcars` that contains 4 cylinder engines. 

7 Get subset of `mtcars` that contains 4 cylinder engines and whose weight is less than 2,000 pounds.

8 Store the subset of `mtcars` that contains 4 cylinder engines as a new data frame named `four_cyl`. Then plot the MPG versus Weight with the following command:

```{r}
#plot(four_cyl$mpg~four_cyl$wt, cex = 1.3, col = "blue", pch = 16, xlab = "Weight (1000 lbs)",ylab = "Miles Per Gallon (Fuel Economy)")
```

### Part 2

9. Look at the structure and help file of the `iris` dataset that's built-in with R. 

```{r}
iris
```


10. What are the variables? How many observations? Plot a histogram of each variable except `species` using the `hist` command,it doesn't have to be fancy. Here's an example: `hist(x, col='red', xlab = "Variable")`. Remember, a histogram plots the frequency of observations in the bin range along the x-axis.

11. Now, use your ability to subset a data frame and make a new data frame for each species. 

```{r}
iris_setosa <- iris %>%   
  filter(Species == "setosa") %>%  
  arrange(Sepal.Length)

iris_setosa
```



12. Next plot a histogram of the sepal length and width for each species. 

13. What's the difference between your histograms in number 12 and number 10?

14. Finally, look at the structure and help file for `iris3`. What's the difference between the `iris` data frame and the `iris3` data frame?



