---
title: "Assignment 3"
author: "Your Name"
date: "Due January 28 by midnight"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment 3 Exercises

```{r}
library(tidyverse)
```


We'll use the `iris` dataset in R for the following exercises.

1. Google what iris's are and how they look; try to understand what petal and sepal lengths/widths are. Using `ggplot`, plot the sepal length verse sepal width in one plot and the petal length and petal width in another plot.

2. Is there an association in either graph? Desribe them. 

3. Filter the `iris` data for the species `virginica` and have it output to the screen. 

4. Filter for the species `virginica` again and then arrange in descending order of sepal length. 

5. The measurements for the iris data are in centimeters. Use the `mutate` verb to create a new column called `slmm`containing the sepal length in millimeters. (Multiply by 10) Repeat this for each of the four original measurements with appropriate column titles. 

```{r}
iris_mod <- iris %>% mutate(slmm = Sepal.Length*10,
                            swmm = Sepal.Width*10,
                            plmm = Petal.Length*10,
                            pwmm = Petal.Width*10)

hist(iris_mod$slmm, col = "light green")
```



6. Plot sepal length verse sepal width in one plot and the petal length and petal width in another plot for the millimeter measurements (similar to #1). Do the relationships change after this unit conversion?


