---
title: "Introduction to Using R: Working with Numbers and Vectors"
author: "Author: Thomas Robacker"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = T)
```

In its simplest role, R functions as an overgrown desktop calculator. In this tutorial we'll discuss how to use the software for arithmetic, storing calculations, learn about vectors, and how to use several useful commands. 

## Arithmetic in R

All basic arithmetic operations are ready for use at the command prompt. You can use the `+,-,*,` and `/` symbols as in a typical hand-held calculator. You can create exponents (or powers) using the `^` symbol, and control the order of the calculations using parentheses, `()`. R follows the standard mathematical rules of parentheses, exponents, multiplication, division, addition,   and subtraction (PEMDAS). 

Here are some examples that you can type into R (you should, to get used to it):

```{r}
3+5
1/3
9/3+1
4^2
```

You can find the square root of any positive number with the `sqrt` function. For example: 
```{r}
sqrt(16)
```

Below are a couple examples of translating more complex arithemtic formulas into R. To express
\[
\frac{1+10+11+14+3+6+9}{7}
\]
in R:
```{r}
(1+10+11+14+3+6+9)/7
```
To express
\[
\frac{5^{3+1/2}\times(6-2)}{61-3+4}
\]
in R:
```{r}
(5^(3+1/2)*(6-2))/(61-3+4)
```

## Objects

So far, R has simply displayed the results of our calculations to the screen. Often times we will want to store values or more complicated things into symbols. If we want to save the results and perform further operations, we need to be able to *assign* the results of a computation to an *object* in the current workspace. This amounts to simply storing an item or result under a given name to be accessed later, without having to do the calculations all over again or multiple times. 

We'll use the terms *assign* and *store* interchangeably. You can specify an assignment in R in two ways: using arrow notation (`<-`) or using a single equal sign (`=`). Let's try both below. We store or assign the value `5` into the object `x` and `7` into `y`. 

```{r}
x<-5 
x
y=7
y
```

Notice that in order to print the result, we must type `x` or `y` to the console to display it. This is nice since we may not explicitly desire to see the output of certain calculations. We can do subsequent operations with these objects. For example, below we calculate the sum of `x` and `y` and store it in `my.number` and print the results:

```{r}
my.number<-x+y
my.number
```

As we get a collection of objects you might wonder which ones you have stored. In R Studio you can see these under the *environment* tab on the right hand side of the default view. You can also type `ls()` into the console. 

```{r}
ls()
```


Although `=` and `<-` do the same thing, it is wise to be consistent. We'll stick with the `<-`, because the potential confusion in using the `=` symbol for mathematical notation. We will reserve the `=` symbol for setting function arguments which we'll see shortly.

You can name objects almost anything as long as the name begins with a letter, avoids symbols (underscores and periods are fine though), and avoids the handful of "reserved" words such those used for special values in R, like the symbol `NA` for a missing value or the word `if` or `for` for a conditional statement or a for-loop.

## Vectors

Often we'll want to perform calculations or comparisons on multiple entities or work with a single group of values instead of many individuals. For simplicity, we'll work only numeric entries right now though many of the utilities we use here apply to non-numeric values too. 

The *vector* is the essential building block for handling multiple items in R. In a numeric sense, a vector is simply a collection of observations or measurements concerning a single variable, for example, the number of coffees an individual drinks daily. More complicated data sets may consist of several vectors. The function/command for creating a vector is the single letter `c` with the entries in parentheses seperated by commas. For example, the command below creates a vector called `v` that contains the numbers 1, 3, 8, and 12.

```{r}
v <- c(1,1,2,3,5,8)
v
```

You also combine vectors. Below we make another vector called `v2` and then make yet another called `v3` which combines both `v` and `v2`:

```{r}
v2<-c(13,21,34)
v3<-c(v,v2)
v3
```

## Sequences, Repetition, Sums, and Lengths

In this section we'll discuss some common and useful functions (or commands) with R vectors. They include the commands `seq`, `rep`, `sum`, and `length`.

### Sequences with `seq`

First, let's create an equally spaced sequence of increasing numeric values. This is useful when plotting data points. The easiest way to create such a sequence is to use the colon operator. For example,
```{r}
3:18
```
created a *vector* of values from 3 to 17 in steps of 1. 

To generalize this a little more we can use the `seq` command. This functions takes the *arguments* of a `from` value, a `to` value, and a `by` value, and it returns the corresponding sequence as a numeric vector. For example, let's make the sequence from 3 to 18 (as before) but in steps of 3:
```{r}
seq(from=3,to=18,by=3)
```
Viola, we have a sequence with intervals of 3 rather than 1. Notice that we had to put `=` symbol to specify the value of each argument. (An *argument* is just an input to a function/command)

### Repetition with `rep`

Sequences are extremely useful, but sometimes we may just want to repeat a certain value. You do this with the `rep` command. For example, let's make a vector that contains fifteen 2's:
```{r}
rep(x=2,times=15)
```
If you wanted to save these results, remember that we *store* it in an *object* that you name:
```{r}
two.vec<-rep(x=2,times=15)
two.vec
```
The `rep` functions is given a single value (or a vector of values) as its argment `x` and a value for the arguments `times` which provides the number of times to repeat `x`.

### Sorting with `sort`

Sorting a vector in increasing or decreasing order of its elements is a simple operation that commonly arises in everyday tasks. The `sort` function takes care of this for us. The commands below create a vector `z` containing a few numeric elements and sorts them in increasing value.
```{r}
z<-c(1,-15,5,-1,0,5,-20,3)
sort(z)
```
If you wanted to sort the vector in decreasing order then you add the argument `decreasing=T`.
```{r}
sort(z,decreasing=T)
```

The `sort` function is pretty straightforward. You supply a vector to the function as the argument `x`, and a second argument, `decreasing`, which indicates the order in which you want to sort. The `decreasing` argument takes a type of value called a *logical* value. A logical value is one of two case-sensitive values: `TRUE` or `FALSE`. You can also just type `T` or `F`. Logicals are often used to indicate the satisfaction or failure of a certain *condition*, and form a integral part to all programming languages. We'll use them as "on/off" instructions for a number of commands/functions. 


Notice that when we typed `sort(z,decreasing=T)` above, we specified that the argument `decreasing` take the value `T` or `TRUE` (We'll usually use the short-to-type values of `T` and `F`). In the command before that we simply typed `sort(z)` and it sorted the values in increasing order. This is because the *default* value of the argument `decreasing` is `FALSE` or `F`, so we didn't have to change it. Many functions contain default values and we'll indicate these as we cover them. 

### Summing with `sum`

If we want to add, or *sum*, up all the elements of a numeric vector then we can use the invaluable `sum` command. You simply insert a vector as the argument.
```{r}
my.vec <- c(1,1,2,3,5,8,13,21)
sum(my.vec)
```

### Finding a Vector Length with `length`

When dealing with data, especially when uploading it from other places, we may want to determine how many entries exist in a vector. The command `length` will do this for us. For example, the length of `my.vec` from the last example is obtained as follows
```{r}
length(my.vec)
```

Let's do a combination of a couple functions we've learned so far. You may recall that the *mean* or *average* of a set of values is the sum of the values divided by how many there are. Yes, there is a built-in function called `mean` for this, but let's just illustrate this using what we know. Let's find the mean for `my.vec`:

```{r}
my.mean <- sum(my.vec)/length(my.vec)
my.mean
```
This is the correct value, you could also check it by typing `mean(my.vec)` in your console. 

### Getting Help with Functions/Commands

Finally, let's see how to get help with functions and commands and knowing what arguments to give them or discovering what their default values may be. To get help on a particular function, simply type `?` before your command. For example, type `?sort` into your console.

You'll see a description of the `sort` command come up under the "help" tab in R Studio. In particular for us now, look at the "Usage" section of the help file. The first line is `sort(x, decreasing = FALSE, ...)`. This shows us the arguments we may pass to the `sort` function. If you scroll down the help file you'll see the "Arguments" section, this tells us what each argument is or could be. So, `x` may be a numeric vector (among many more) which is how we have used it so far. Noting the `decreasing = FALSE` under "Usage" tells us that's the default value for the argument `decreasing`, which is why we only had to type `sort(z)` earlier to get increasing order. The ellipsis `...` indicates that we may pass additional arguments as described in the "Arguments" section of the help file.

You don't need to memorize these arguments or commands, as you use them you'll get used to them and how they operate. We will go over every command/function we use in this course in the detail we require and you can always get help with the `?` operator for a command. Don't forget the Golden Rule!

### Exercises (Assignment 1)

In a new R Markdown file, I would like you to copy and complete the following exercises. You must show all of your code and output. 

1. Create and store a sequence of values from 1 to 50 that progresses in steps of 0.5.

```{r}
sequence <- seq(from = 1, to = 50, by = 0.5)
sequence
```



2. Overwrite the object from number 1 with the same sequence with the order reversed. 

3. Create a vector labeled `my.vec` that contains the numbers 3,1,10,24,31,70, and 110. Using the `sum`, `length`, and `sort` commands compute the sum, the length, and sort the vector my.vec in increasing and decreasing order.

4.	Use the `sum` and `length` functions to calculate the mean of `my.vec` then use the `mean` function to compute the mean. Can you figure out how to find the median, standard deviation, and variance? Show me (Use the internet!)

5. Once you've found the median, standard deviation, and variance of `my.vec`, tell me what those represent. That is, interpret what each of those values means about `my.vec`.

6. The conversion from a temperature measurement in degrees Fahrenheit *F* to Celsius *C* is performed using the following equation:
\[
C=\frac{5}{9}(F-32)
\]
Use vector-oriented behavior to convert the temperatures 45, 77, 20, 19, 101, 120, and 212 in degrees Fahrenheit to degrees Celsius.

```{r}
F <- c(45,77,20,19,101,120,212)
C <- 5/9*(F-32)
C
```



#### References

*The Book of R: A First Course in Programming and Statistics* , Tilman Davies. No Starch Press, Inc. 2016.
